---
title: "Top 100 Recruit RDA Files"
author: "BSA Basketball Group 25-26"
date: "`r format(Sys.Date(), '%m/%d/%y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ONLY NEED TO RUN THIS FILE ONCE

# The Rankings
# Reads the ESPN Top 100 recruiting rankings (class of 2016-2024), cleans 
# the data, and adds tier, conference, and school file labels

```{r}
library(tidyverse)

# Read the rankings data
rankings_raw <- read_csv(
  "2016_to_2024_class_rankings.csv",
  col_names = c("CLASS", "RK", "NAME", "RATING",
                "COLLEGE", "POS", "HT", "HS"),
  show_col_types = FALSE
)

# Standardize basic fields
rankings_base <- rankings_raw %>%
  mutate(
    CLASS        = as.integer(CLASS),
    RK           = as.integer(RK),
    player_clean = str_to_lower(str_trim(NAME))
  )

# Assign each recruit to a ranking tier (1-25, 26-50, 51-75, 76-100)
rankings_tier <- rankings_base %>%
  mutate(
    tier = case_when(
      RK >= 1  & RK <= 25  ~ "1-25",
      RK >= 26 & RK <= 50  ~ "26-50",
      RK >= 51 & RK <= 75  ~ "51-75",
      RK >= 76 & RK <= 100 ~ "76-100",
      TRUE                 ~ NA_character_
    )
  )

# Some ESPN school names don't match Sports Reference's file naming
# This is a manual lookup table
# "ESPN college name" = "Sports reference file prefix"
school_mapping_custom <- c(
  "UConn"          = "connecticut",
  "USC"            = "southern-california",
  "Ole Miss"       = "mississippi",
  "Texas A&M"      = "texas-am",
  "NC State"       = "north-carolina-state",
  "St. John's"     = "st-johns-ny"
)

# Apply the manual mapping
# Then auto-generate the file prefix
rankings_school <- rankings_tier %>%
  mutate(
    school_file = school_mapping_custom[COLLEGE],
    school_file = if_else(
      is.na(school_file),
      str_to_lower(str_replace_all(COLLEGE, " ", "-")),
      school_file
    ),
    school_file = str_replace_all(school_file, "[^a-z0-9-]", "")
  )

# Assign each school to a conference (Power 4 or Other)
rankings_conf <- rankings_school %>%
  mutate(
    conference = case_when(
      COLLEGE %in% c("Alabama","Arkansas","Auburn","Florida","Georgia",
                     "Kentucky","LSU","Mississippi State","Ole Miss","Missouri",
                     "South Carolina","Tennessee","Texas","Texas A&M",
                     "Vanderbilt","Oklahoma") ~ "SEC",
      COLLEGE %in% c("Illinois","Indiana","Iowa","Maryland","Michigan",
                     "Michigan State","Minnesota","Nebraska","Northwestern",
                     "Ohio State","Penn State","Purdue","Rutgers","UCLA",
                     "USC","Oregon","Washington","Wisconsin") ~ "Big Ten",
      COLLEGE %in% c("Boston College","Clemson","Duke","Florida State",
                     "Georgia Tech","Louisville","Miami","NC State",
                     "North Carolina","Notre Dame","Pittsburgh","Syracuse",
                     "Virginia","Virginia Tech","Wake Forest",
                     "California","Stanford","SMU") ~ "ACC",
      COLLEGE %in% c("Arizona","Arizona State","Baylor","BYU","Cincinnati",
                     "Colorado","Houston","Iowa State","Kansas","Kansas State",
                     "Oklahoma State","TCU","Texas Tech","UCF","Utah",
                     "West Virginia") ~ "Big 12",
      TRUE ~ "Other"
    )
  )

rankings_pos <- rankings_conf

# Cleaned recruiting rankings with tiers + conferences
save(rankings_pos, file = "rankings_pos.rda")

```

# Advanced Stats
# Reads all Sports Reference advanced stat CSVs for every school
# across all seasons (2017-2025), and stacks them into one table
# Also maps position labels (G, W, F, P, C)

```{r}
library(stringr)
library(purrr)

# Root folder that contains subfolders for each year
# Each subfolder is named: YYYY_advanced_data
adv_root <- "C:/Users/joshu/OneDrive/Documents/UCLA/Clubs/BSA/Freshman Recruiting" # Change file path

years <- 2017:2025

# Loop over each year, final all CSV files, read them all
advanced_all <- map_df(years, function(y) {
  folder <- file.path(adv_root, paste0(y, "_advanced_data"))
  files  <- list.files(folder, pattern = "_wbb_advanced_\\d{4}\\.csv$",
                       full.names = TRUE)

  map_df(files, function(f) {
    read_csv(f, show_col_types = FALSE) %>%
      mutate(source_file = basename(f))
  })
}) %>%
  mutate(
    player_clean = str_to_lower(str_trim(player_name)),
    school_clean = str_to_lower(str_trim(school)),
    season_start = as.integer(substr(season, 1, 4))
  )

# Map Sports Reference raw position labels to 5 position groups
# G = Guard, W = Wing (G-F or F-G), F = Forward, P = Post (F-C or C-F)
advanced_all <- advanced_all %>%
  mutate(
    pos_raw = str_to_upper(str_trim(pos)),
    pos_group = case_when(
      str_detect(pos_raw, "G-F") ~ "W",
      str_detect(pos_raw, "F-G") ~ "W",
      str_detect(pos_raw, "F-C") ~ "P",
      str_detect(pos_raw, "C-F") ~ "P",
      str_detect(pos_raw, "G")   ~ "G",
      str_detect(pos_raw, "F")   ~ "F",
      str_detect(pos_raw, "C")   ~ "C",
      TRUE                       ~ NA_character_
    )
  )

# All school advanced stats, 2017–2025
save(advanced_all, file = "advanced_all.rda")

```

# Linked Tables
# Join rankings to each player's freshman season stats

```{r}
load("rankings_pos.rda")
load("advanced_all.rda")

# Find each player's first season in the advanced data
freshman_season <- advanced_all %>%
  group_by(player_clean) %>%
  summarise(freshman_season_start = min(season_start, na.rm = TRUE),
            .groups = "drop")

# Filter advanced_all to only each player's freshman season rows
advanced_freshman <- advanced_all %>%
  inner_join(freshman_season,
             by = c("player_clean", "season_start" = "freshman_season_start"))

# Join the rankings to freshman-year stats
# Match on player name and school
stats_joined <- rankings_pos %>%
  select(CLASS, RK, NAME, COLLEGE, conference,
         tier, player_clean, school_file) %>%
  left_join(
    advanced_freshman,
    by = c("player_clean",
           "school_file" = "school_clean")
  )

# Rankings joined to each player's freshman-year stats
save(stats_joined, file = "stats_joined.rda")

```

